---
title: "betydata"
subtitle: "Public Data from the BETYdb Database"
---

```{r}
#| echo: false
library(betydata)
library(dplyr)
```

**betydata** is an R data package providing offline access to public data from [BETYdb](https://betydb.org) (Biofuel Ecophysiological Traits and Yields database). It enables reproducible analyses of plant traits, crop yields, and supporting metadata without requiring database connectivity.

## At a Glance

```{r}
#| echo: false
#| label: tbl-overview
#| tbl-cap: "Dataset summary"
tibble::tibble(
  Metric = c(
    "Total observations", "Unique traits", "Plant species",
    "Research sites", "Literature citations", "Tables"
  ),
  Value = c(
    format(nrow(traitsview), big.mark = ","),
    n_distinct(traitsview$trait),
    format(nrow(species), big.mark = ","),
    nrow(sites),
    nrow(citations),
    16
  )
) |> knitr::kable()
```

## Installation

```r
# install.packages("remotes")
remotes::install_github("PecanProject/betydata")
```

## Quick Start

```{r}
library(betydata)
library(dplyr)

# The primary table -- key columns first, IDs last
traitsview

# Bioenergy crop observations
bioenergy_genera <- c("Miscanthus", "Panicum", "Populus", "Salix", "Saccharum")
traitsview |>
  filter(genus %in% bioenergy_genera) |>
  count(genus, sort = TRUE)
```

## Guides

| Guide | Description |
|-------|-------------|
| [Getting Started](vignettes/getting_started.qmd) | Package overview, data model, and key concepts |
| [Common Analyses](vignettes/common_analyses.qmd) | Practical analysis patterns with dplyr |
| [PFTs & Priors](vignettes/pfts-priors.qmd) | Plant Functional Types and Bayesian prior distributions |
| [Manuscript Reproduction](vignettes/manuscript.qmd) | Reproduce key analyses from LeBauer et al. (2018) |

## Data License

Data: [ODC-By-1.0](https://opendatacommons.org/licenses/by/1-0/)
Code: [BSD-3-Clause](https://opensource.org/licenses/BSD-3-Clause)

## Citation

LeBauer, D. S., et al. (2018). BETYdb: a yield, trait, and ecosystem service database applied to second-generation bioenergy feedstock production. *GCB Bioenergy*. [doi:10.1111/gcbb.12420](https://doi.org/10.1111/gcbb.12420)
