---
title: "Common Analyses with betydata"
vignette: >
  %\VignetteIndexEntry{Common Analyses with betydata}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(eval = requireNamespace("betydata", quietly = TRUE))
```

::: {.callout-note}
## What you will learn

- How to extract and summarize yield data for specific genera
- How to link management practices (fertilization, planting) to yield observations
- Patterns for site-level aggregation, author-based queries, and variable lookups
:::

## Setup

```{r}
library(betydata)
library(dplyr)
```

## Extracting Yield Data for a Genus {#sec-yields}

A common starting point is pulling yield observations for a particular genus and summarizing them. The `Ayield` trait represents above-ground annual yield in Mg/ha.

```{r}
miscanthus_yields <- traitsview |>
  filter(
    genus == "Miscanthus",
    trait == "Ayield"
  ) |>
  select(id, mean, date, sitename, scientificname)

miscanthus_yields

nrow(miscanthus_yields)
```

::: {.callout-tip}
## Tibble Printing

All tables are tibbles, which display the first 10 rows by default. With key columns ordered first (`trait`, `mean`, `units`, `scientificname`, `genus`), the default output is immediately informative without needing `head()` or column subsetting.
:::

## Working with Management Practices {#sec-management}

Management practices (planting dates, fertilization rates, harvest methods) are stored in the `managements` table and linked to experimental treatments through the `managements_treatments` junction table. This linkage connects management details to yield observations in `traitsview`.

```{r}
mgmt_treat <- managements_treatments |>
  left_join(
    managements |> select(id, mgmttype, level, units, date),
    by = c("management_id" = "id")
  )

grass_yields <- traitsview |>
  filter(
    genus %in% c("Miscanthus", "Panicum"),
    trait == "Ayield"
  ) |>
  left_join(mgmt_treat, by = "treatment_id", relationship = "many-to-many")

grass_yields |>
  filter(!is.na(mgmttype)) |>
  count(genus, mgmttype, sort = TRUE)
```

## Nitrogen Fertilization Rates {#sec-nitrogen}

Extracting nitrogen application rates and joining them with yield data enables exploration of yield--nitrogen relationships. Nitrogen management is recorded as `fertilizer_N` or `fertilizer_N_rate` in the `mgmttype` column.

```{r}
nitrogen_rates <- managements |>
  filter(mgmttype %in% c("fertilizer_N", "fertilizer_N_rate")) |>
  left_join(
    managements_treatments |> select(management_id, treatment_id),
    by = c("id" = "management_id")
  ) |>
  select(treatment_id, nrate = level, units)

yields_with_n <- traitsview |>
  filter(
    trait == "Ayield",
    genus %in% c("Miscanthus", "Panicum")
  ) |>
  left_join(nitrogen_rates, by = "treatment_id", relationship = "many-to-many")

yields_with_n |>
  filter(!is.na(nrate)) |>
  summarise(
    n = n(),
    mean_N = round(mean(nrate, na.rm = TRUE), 1),
    mean_yield = round(mean(mean, na.rm = TRUE), 1),
    .by = genus
  ) |>
  knitr::kable(col.names = c("Genus", "N obs", "Mean N rate", "Mean Yield (Mg/ha)"))
```

## Site-Level Aggregation {#sec-sites}

Aggregating trait data by site is useful for spatial analysis and mapping data density across research locations.

```{r}
#| label: tbl-site-summary
#| tbl-cap: "Top research sites by number of records"
site_summary <- traitsview |>
  filter(!is.na(lat), !is.na(lon)) |>
  summarise(
    n_records = n(),
    n_traits = n_distinct(trait),
    n_species = n_distinct(species_id),
    .by = c(site_id, sitename, lat, lon)
  )

site_summary |>
  arrange(desc(n_records)) |>
  head(15) |>
  knitr::kable()
```

::: {.callout-tip}
## Geographic Data

All sites with coordinates have `lat` and `lon` columns in both `traitsview` and the `sites` table. The `sites` table additionally contains `mat` (mean annual temperature) and `map` (mean annual precipitation) for sites where climate data is available.
:::

## Finding Data by Author {#sec-author}

```{r}
lebauer_data <- traitsview |>
  filter(grepl("LeBauer", author, ignore.case = TRUE))

lebauer_data |>
  count(trait, author, citation_year, sort = TRUE)
```

## Most Data-Rich Citations {#sec-citations}

```{r}
#| label: tbl-citations
#| tbl-cap: "Top 10 citations by number of records"
traitsview |>
  count(citation_id, author, citation_year, sort = TRUE) |>
  head(10) |>
  knitr::kable()
```

## Variable and Trait Lookups {#sec-variables}

The `variables` table provides units, descriptions, and valid ranges for each measured trait. This is useful for understanding what a trait measures and checking whether observed values are within expected bounds.

```{r}
variables |>
  filter(name %in% c("SLA", "Vcmax", "leaf_respiration_rate_m2", "Ayield")) |>
  select(name, units, description, min, max) |>
  knitr::kable()
```

## Performance {#sec-performance}

Since all tables are loaded in memory as R data frames, filtering and joining operations run at in-memory speed with no network overhead.

```{r}
system.time({
  result <- traitsview |>
    filter(
      genus %in% c("Miscanthus", "Panicum", "Populus"),
      trait %in% c("SLA", "Vcmax", "Ayield"),
      checked == 1
    ) |>
    summarise(
      n = n(),
      mean = mean(mean, na.rm = TRUE),
      .by = c(genus, trait)
    )
})
```

## References

- LeBauer, D. S., et al. (2018). BETYdb: a yield, trait, and ecosystem service database applied to second-generation bioenergy feedstock production. *GCB Bioenergy*. [doi:10.1111/gcbb.12420](https://doi.org/10.1111/gcbb.12420)
