---
title: "Reproducing BETYdb Manuscript Analyses"
vignette: >
  %\VignetteIndexEntry{Reproducing BETYdb Manuscript Analyses}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

::: {.callout-note}
## What you will learn

- How to reproduce key figures and tables from the BETYdb publication
- How to summarize trait and yield data by genus, trait, and site
- How current data compares to the 2017 snapshot used in the paper
:::

## Introduction

This vignette reproduces key analyses from the BETYdb manuscript (LeBauer et al., 2018) using the `betydata` package.

**Citation:** LeBauer, D. S., et al. (2018). BETYdb: a yield, trait, and ecosystem service database applied to second-generation bioenergy feedstock production. *GCB Bioenergy*. [doi:10.1111/gcbb.12420](https://doi.org/10.1111/gcbb.12420)

## Setup

```{r}
library(betydata)
library(dplyr)
library(ggplot2)

theme_set(theme_bw(base_size = 10, base_family = "sans"))
```

## Figure 1: Data Summary by Genus {#sec-fig1}

The manuscript presents trait and yield counts for key bioenergy genera. We reproduce this using `traitsview`:

```{r}
#| label: tbl-genus-summary
#| tbl-cap: "Trait and yield counts by bioenergy genus"
bioenergy_genera <- c("Miscanthus", "Panicum", "Populus", "Saccharum",
                       "Pinus", "Salix", "Robinia")

genus_summary <- traitsview |>
  filter(genus %in% bioenergy_genera, checked >= 0) |>
  summarise(
    n_traits = sum(result_type == "traits", na.rm = TRUE),
    n_yields = sum(result_type == "yields", na.rm = TRUE),
    total = n(),
    .by = genus
  ) |>
  arrange(desc(total))

genus_summary |>
  knitr::kable(col.names = c("Genus", "Traits", "Yields", "Total"))
```

::: {.callout-important}
## Comparison Notes

Counts may differ from the published manuscript because:

1. **QA/QC filtering:** betydata excludes `checked = -1` (failed QA/QC records)
2. **Snapshot date:** betydata was exported from a current database snapshot; the manuscript used 2017 data
3. **Access level:** betydata includes only public data from BETYdb
:::

## Figure 2: Trait Records by Genus {#sec-fig2}

```{r}
#| label: fig-trait-by-genus
#| fig-cap: "Number of records for focal traits across bioenergy genera (log scale)"
#| fig-height: 6
focal_traits <- c("Ayield", "leafN", "LAI", "SLA", "Vcmax",
                   "leaf_respiration_rate_m2", "Jmax")

trait_counts <- traitsview |>
  filter(
    genus %in% bioenergy_genera,
    trait %in% focal_traits,
    checked >= 0
  ) |>
  count(genus, trait, name = "n")

ggplot(trait_counts, aes(x = genus, y = n, fill = trait)) +
  geom_col(position = "dodge") +
  scale_y_log10(breaks = c(1, 10, 100, 1000, 10000)) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Number of Records (log scale)",
    fill = "Trait"
  ) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

## Figure 3: Trait Distributions {#sec-fig3}

The manuscript displays histograms of trait values across genera, showing the spread of measured values for key ecophysiological parameters.

```{r}
#| label: fig-trait-distributions
#| fig-cap: "Distribution of trait values for bioenergy genera"
#| fig-height: 8
hist_traits <- c("Ayield", "SLA", "Vcmax", "LAI")

trait_data <- traitsview |>
  filter(
    trait %in% hist_traits,
    !is.na(mean),
    checked >= 0,
    genus %in% bioenergy_genera
  )

ggplot(trait_data, aes(x = mean, fill = genus)) +
  geom_histogram(bins = 25, alpha = 0.7) +
  facet_wrap(~trait, scales = "free", ncol = 2) +
  labs(
    x = "Observed Value",
    y = "Count",
    fill = "Genus"
  ) +
  theme(
    legend.position = "bottom",
    strip.background = element_blank()
  )
```

## Table 1: Database Contents Summary {#sec-table1}

```{r}
#| label: tbl-contents
#| tbl-cap: "Data records by genus (top 15 with >= 100 records)"
contents <- traitsview |>
  filter(checked >= 0) |>
  summarise(
    n_traits = sum(result_type == "traits", na.rm = TRUE),
    n_yields = sum(result_type == "yields", na.rm = TRUE),
    total = n(),
    .by = genus
  ) |>
  filter(total >= 100) |>
  arrange(desc(total))

knitr::kable(
  head(contents, 15),
  col.names = c("Genus", "Traits", "Yields", "Total")
)
```

## Yield Meta-Analysis Subset {#sec-meta-analysis}

The manuscript includes a meta-analysis of *Miscanthus* and switchgrass (*Panicum*) yields. Here we extract the relevant subset and compute summary statistics.

```{r}
#| label: tbl-yield-ma
#| tbl-cap: "Yield meta-analysis summary for Miscanthus and Panicum"
yield_ma <- traitsview |>
  filter(
    genus %in% c("Miscanthus", "Panicum"),
    trait == "Ayield",
    !is.na(lat),
    !is.na(lon),
    !is.na(mean),
    checked >= 0
  ) |>
  select(
    id, genus, scientificname, mean, units,
    n, stat, statname, lat, lon,
    author, citation_year, sitename, site_id
  )

yield_ma |>
  summarise(
    n_records = n(),
    mean_yield = round(mean(mean), 1),
    sd_yield = round(sd(mean), 1),
    n_sites = n_distinct(site_id),
    .by = genus
  ) |>
  knitr::kable(col.names = c("Genus", "Records", "Mean Yield (Mg/ha)", "SD", "Sites"))
```

## Geographic Distribution {#sec-geography}

```{r}
#| label: fig-geo-distribution
#| fig-cap: "Geographic distribution of Miscanthus and switchgrass yield observations"
#| fig-width: 8
#| fig-height: 5
ggplot(yield_ma, aes(x = lon, y = lat, color = genus)) +
  geom_point(alpha = 0.6, size = 2) +
  borders("world", colour = "grey70", fill = NA) +
  coord_quickmap(xlim = c(-130, 50), ylim = c(20, 70)) +
  labs(
    x = "Longitude",
    y = "Latitude",
    color = "Genus"
  ) +
  theme_minimal(base_size = 12)
```

::: {.callout-tip}
## Extending This Analysis

The `sites` table contains additional geographic and climate metadata (`mat`, `map`, `soil`) that can be joined to enable climate-response analyses.
:::

## References

- LeBauer, D. S., et al. (2018). BETYdb: a yield, trait, and ecosystem service database applied to second-generation bioenergy feedstock production. *GCB Bioenergy*. [doi:10.1111/gcbb.12420](https://doi.org/10.1111/gcbb.12420)
