---
title: "Reproducing BETYdb Manuscript Analyses"
subtitle: "Offline Replication using betydata"
author: "Akash B V"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing BETYdb Manuscript Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

This vignette reproduces key analyses from the BETYdb manuscript (LeBauer et al., 2018) using the offline `betydata` package. The original analyses queried a live PostgreSQL database; here we demonstrate equivalent results using packaged data.

**Citation:** LeBauer, D. S., et al. (2018). BETYdb: a yield, trait, and ecosystem service database applied to second-generation bioenergy feedstock production. *GCB Bioenergy*. https://doi.org/10.1111/gcbb.12420

## Setup
```{r load-packages}
library(betydata)
library(dplyr)
library(ggplot2)

# Set theme
theme_set(theme_bw(base_size = 10, base_family = "sans"))
```

## Figure 1: Data Summary by Genus

The manuscript presents trait and yield counts for bioenergy genera.
```{r genus-summary}
data(traitsview)

# Define focal genera (from manuscript)
bioenergy_genera <- c("Miscanthus", "Panicum", "Populus", "Saccharum", 
                      "Pinus", "Salix", "Robinia")

# Compute counts by genus and result type
genus_summary <- traitsview |>
  filter(genus %in% bioenergy_genera, checked >= 0) |>
  group_by(genus) |>
  summarise(
    n_traits = sum(result_type == "traits", na.rm = TRUE),
    n_yields = sum(result_type == "yields", na.rm = TRUE),
    total = n(),
    .groups = "drop"
  ) |>
  arrange(desc(total))

genus_summary
```

### Comparison Notes

The counts differ slightly from the published manuscript because:

1. betydata excludes `checked = -1` (failed QA/QC records)
2. Snapshot date: betydata was exported on `r format(Sys.Date(), "%Y-%m-%d")`; the manuscript used 2017 data
3. Access level filtering: betydata includes only public data (`access_level < 4`)

## Figure 2: Trait Records by Genus
```{r trait-counts-plot, fig.height = 6}
# Key traits analyzed in manuscript
focal_traits <- c("Ayield", "leafN", "LAI", "SLA", "Vcmax", 
                  "leaf_respiration_rate_m2", "Jmax")

trait_counts <- traitsview |>
  filter(
    genus %in% bioenergy_genera,
    trait %in% focal_traits,
    checked >= 0
  ) |>
  count(genus, trait, name = "n")

ggplot(trait_counts, aes(x = genus, y = n, fill = trait)) +
  geom_col(position = "dodge") +
  scale_y_log10(breaks = c(1, 10, 100, 1000, 10000)) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Number of Records (log scale)",
    fill = "Trait"
  ) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

## Figure 3: Trait Distributions

The manuscript displays histograms of trait values across genera.
```{r trait-distributions, fig.height = 8}
# Select key traits for visualization
hist_traits <- c("Ayield", "SLA", "Vcmax", "LAI")

trait_data <- traitsview |>
  filter(
    trait %in% hist_traits,
    !is.na(mean),
    checked >= 0,
    genus %in% c(bioenergy_genera, "Other")
  ) |>
  mutate(
    genus = if_else(genus %in% bioenergy_genera, genus, "Other"),
    genus = factor(genus)
  )

ggplot(trait_data, aes(x = mean, fill = genus)) +
  geom_histogram(bins = 25, alpha = 0.7) +
  facet_wrap(~trait, scales = "free", ncol = 2) +
  labs(
    x = "Observed Value",
    y = "Count",
    fill = "Genus"
  ) +
  theme(
    legend.position = "bottom",
    strip.background = element_blank()
  )
```

## Table 1: Database Contents Summary
```{r contents-table}
contents <- traitsview |>
  filter(checked >= 0) |>
  group_by(genus) |>
  summarise(
    n_traits = sum(result_type == "traits", na.rm = TRUE),
    n_yields = sum(result_type == "yields", na.rm = TRUE),
    total = n(),
    .groups = "drop"
  ) |>
  filter(total >= 100) |>  # Genera with substantial data
  arrange(desc(total))

# Top 15 genera
knitr::kable(
  head(contents, 15),
  col.names = c("Genus", "Traits", "Yields", "Total"),
  caption = "Data records by genus (top 15)"
)
```

## Yield Meta-Analysis Subset

The manuscript includes a meta-analysis of Miscanthus and Switchgrass yields. Here we extract the relevant subset:
```{r yield-meta-analysis}
# following manuscript criteria:
# - Miscanthus and Panicum only
# - yield trait (Ayield)
# - with site coordinates

yield_ma <- traitsview |>
  filter(
    genus %in% c("Miscanthus", "Panicum"),
    trait == "Ayield",
    !is.na(lat),
    !is.na(lon),
    !is.na(mean),
    checked >= 0
  ) |>
  select(
    id, genus, scientificname, mean, units, 
    n, stat, statname, lat, lon,
    author, citation_year, sitename, site_id
  )

yield_ma |>
  group_by(genus) |>
  summarise(
    n_records = n(),
    mean_yield = mean(mean),
    sd_yield = sd(mean),
    n_sites = n_distinct(site_id),
    .groups = "drop"
  )
```

## Geographic Distribution
```{r geographic-map, fig.width = 8, fig.height = 5}
# Simple map of yield observation locations
ggplot(yield_ma, aes(x = lon, y = lat, color = genus)) +
  geom_point(alpha = 0.6, size = 2) +
  borders("world", colour = "grey70", fill = NA) +
  coord_quickmap(xlim = c(-130, 50), ylim = c(20, 70)) +
  labs(
    x = "Longitude",
    y = "Latitude",
    color = "Genus",
    title = "Miscanthus and Switchgrass Yield Observations"
  ) +
  theme_minimal()
```

## Session Info
```{r session-info}
sessionInfo()
```