---
title: "Getting Started with betydata"
subtitle: "Offline Access to BETYdb Plant Trait Data"
author: "Akash B V"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with betydata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The `betydata` package provides offline access to public data from BETYdb, the Biofuel Ecophysiological Traits and Yields database. This vignette demonstrates the basics of loading and exploring the packaged datasets.

**Key features:**

-   No database connection required
-   Versioned, reproducible data snapshots
-   Pre-joined, analysis-ready data frames
-   Compatible with tidyverse workflows

## Loading the Package

```{r load-package}
library(betydata)
```

## Available Datasets

The package exports 16 datasets. List them all:

```{r list-datasets}
data(package = "betydata")$results[, c("Item", "Title")]
```

## The Primary Dataset: traitsview

The `traitsview` dataset is a denormalized view combining traits and yields with their associated metadata:

```{r explore-traitsview}
data(traitsview)

# Dimensions
dim(traitsview)

# Column names
names(traitsview)
```

### Key Columns

| Column           | Description                                   | Type      |
|----------------|-----------------------------------------|----------------|
| `trait`          | Variable name (e.g. "SLA", "Vcmax", "Ayield") | character |
| `mean`           | Observed value                                | numeric   |
| `units`          | Measurement units                             | character |
| `scientificname` | Species scientific name                       | character |
| `sitename`       | Research site name                            | character |
| `author`         | Citation author                               | character |
| `checked`        | QA/QC status (0 = unchecked, 1 = verified)    | integer   |

## Basic Exploration

```{r basic-exploration}
# Preview
head(traitsview[, c("trait", "mean", "units", "scientificname", "author")])

# Most common traits
if (requireNamespace("dplyr", quietly = TRUE)) {
  library(dplyr)
  traitsview |>
    count(trait, sort = TRUE) |>
    head(15)
}
```

## Filtering by QA/QC Status

```{r qa-filtering}
# Distribution of checked values
table(traitsview$checked, useNA = "ifany")

# Work with verified records only
verified <- traitsview[traitsview$checked == 1, ]
nrow(verified)
```

## Support Tables

### Species Taxonomy

```{r species}
data(species)
dim(species)
head(species[, c("id", "scientificname", "genus", "commonname")])
```

### Variables (Trait Definitions)

```{r variables}
data(variables)

# Common model parameters
variables[variables$name %in% c("SLA", "Vcmax", "leaf_respiration_rate_m2"), 
          c("name", "units", "description")]
```

### Sites

```{r sites}
data(sites)

# Sites with climate data
sites_with_climate <- sites[!is.na(sites$mat) & !is.na(sites$map), ]
nrow(sites_with_climate)
```

## Example: Bioenergy Crop Yields

```{r bioenergy-yields}
library(dplyr)

# Define bioenergy genera
bioenergy_genera <- c("Miscanthus", "Panicum", "Populus", "Salix", "Saccharum")

# Extract yield data
yields <- traitsview |>
  filter(
    trait == "Ayield",
    genus %in% bioenergy_genera,
    !is.na(mean)
  ) |>
  select(genus, mean, units, sitename, author, citation_year, lat, lon)

# Summary by genus
yields |>
  group_by(genus) |>
  summarise(
    n = n(),
    mean_yield = mean(mean, na.rm = TRUE),
    sd_yield = sd(mean, na.rm = TRUE),
    .groups = "drop"
  )
```

## Working with Plant Functional Types (PFTs)

```{r pfts-example}
data(pfts)
data(pfts_species)

# Find PFTs for common bioenergy species
data(species)
miscanthus_sp <- species |> 
  filter(genus == "Miscanthus") |>
  pull(id)

# Which PFTs include Miscanthus species?
pfts_species |>
  filter(specie_id %in% miscanthus_sp) |>
  left_join(pfts |> select(id, name), by = c("pft_id" = "id")) |>
  distinct(name)
```

## Next Steps

See additional vignettes for:

-   `vignette("manuscript")` — Reproduce key figures from LeBauer et al. (2018)
-   `vignette("sql-analogs")` — Offline equivalents of common SQL queries
-   `vignette("pfts-priors")` — Working with PFTs and Bayesian priors

## Session Info

```{r session-info}
sessionInfo()
```
