---
title: "Working with Plant Functional Types and Priors"
subtitle: "Bayesian Model Parameterization using betydata"
author: "Akash B V"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Plant Functional Types and Priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates how to work with Plant Functional Types (PFTs) and Bayesian prior distributions using `betydata`. These are central to the PEcAn ecosystem modeling workflow.

Key concepts:

* **PFT (Plant Functional Type)**: A grouping of species with similar functional characteristics used for ecosystem modeling
* **Prior**: A probability distribution encoding expert knowledge about a parameter before observing data
* **Posterior**: Updated distribution after combining prior with observations

## Setup
```{r}
library(betydata)
library(dplyr)
library(ggplot2)

# Load PFT and prior datasets
data(pfts)
data(priors)
data(variables)
data(pfts_species)
data(pfts_priors)
data(species)
data(traitsview)
```

## Understanding PFTs

### Available PFTs
```{r}
# Number of PFTs
nrow(pfts)

# PFT structure
glimpse(pfts)

# Common PFT names
pfts |>
  count(name, sort = TRUE) |>
  head(20)
```

### Species Membership

Each PFT contains species assigned by researchers:
```{r}
# How many species per PFT?
species_per_pft <- pfts_species |>
  count(pft_id, name = "n_species")

summary(species_per_pft$n_species)

# Find species in a specific PFT (e.g. Miscanthus)
miscanthus_pft <- pfts |> filter(grepl("miscanthus", name, ignore.case = TRUE))

if (nrow(miscanthus_pft) > 0) {
  miscanthus_species <- pfts_species |>
    filter(pft_id %in% miscanthus_pft$id) |>
    left_join(species |> select(id, scientificname), by = c("specie_id" = "id"))
  
  print(miscanthus_species)
}
```

## Working with Priors

### Prior Distribution Structure
```{r}
glimpse(priors)

# Distribution types used
priors |> count(distn, sort = TRUE)

# Variables with priors
priors |>
  left_join(variables |> select(id, name), by = c("variable_id" = "id")) |>
  count(name, sort = TRUE) |>
  head(15)
```

### Visualizing a Prior Distribution
```{r}
# Get SLA priors first
sla_var_id <- variables |> 
  filter(name == "SLA") |> 
  pull(id)

sla_priors <- priors |>
  filter(variable_id %in% sla_var_id)

# Function to evaluate any prior distribution
eval_prior_density <- function(p, x = NULL) {
  # Generate appropriate x range if not provided
  if (is.null(x)) {
    x <- switch(p$distn,
      "gamma" = seq(0, qgamma(0.999, p$parama, p$paramb), length.out = 1000),
      "norm" = seq(p$parama - 4*p$paramb, p$parama + 4*p$paramb, length.out = 1000),
      "lnorm" = seq(0.001, qlnorm(0.999, p$parama, p$paramb), length.out = 1000),
      "beta" = seq(0.001, 0.999, length.out = 1000),
      "unif" = seq(p$parama, p$paramb, length.out = 1000),
      "weibull" = seq(0, qweibull(0.999, p$parama, p$paramb), length.out = 1000),
      "exp" = seq(0, qexp(0.999, p$parama), length.out = 1000),
      seq(0, 100, length.out = 1000)
    )
  }
  
  y <- switch(p$distn,
    "gamma"   = dgamma(x, shape = p$parama, rate = p$paramb),
    "norm"    = dnorm(x, mean = p$parama, sd = p$paramb),
    "lnorm"   = dlnorm(x, meanlog = p$parama, sdlog = p$paramb),
    "beta"    = dbeta(x, shape1 = p$parama, shape2 = p$paramb),
    "unif"    = dunif(x, min = p$parama, max = p$paramb),
    "weibull" = dweibull(x, shape = p$parama, scale = p$paramb),
    "exp"     = dexp(x, rate = p$parama),
    rep(0, length(x))  # fallback
  )
  
  list(x = x, y = y)
}

# Plot if SLA priors exist
if (nrow(sla_priors) > 0) {
  p <- sla_priors[1, ]
  result <- eval_prior_density(p)
  
  ggplot(data.frame(x = result$x, y = result$y), aes(x, y)) +
    geom_line(linewidth = 1) +
    labs(
      title = "Prior distribution for SLA",
      subtitle = paste0(p$distn, "(", round(p$parama, 2), ", ", round(p$paramb, 2), ")"),
      x = "SLA (m2/kg)",
      y = "Density"
    ) +
    theme_minimal()
}
```

## Linking PFTs to Priors

The `pfts_priors` table connects PFTs to their associated prior distributions:
```{r}
# Structure of the mapping
glimpse(pfts_priors)

# Priors per PFT
priors_per_pft <- pfts_priors |>
  count(pft_id, name = "n_priors")

summary(priors_per_pft$n_priors)

# Example: Find all priors for a specific PFT
if (exists("miscanthus_pft") && nrow(miscanthus_pft) > 0) {
  pft_id <- miscanthus_pft$id[1]
  
  pft_priors <- pfts_priors |>
    filter(pft_id == !!pft_id) |>
    left_join(priors, by = c("prior_id" = "id")) |>
    left_join(variables |> select(id, name, units), by = c("variable_id" = "id"))
  
  pft_priors |>
    select(name, units, distn, parama, paramb)
}
```

## Comparing Prior to Data

A key use case: compare the prior distribution to observed data.
```{r}
# SLA observations for grasses
sla_data <- traitsview |>
  filter(
    trait == "SLA",
    genus %in% c("Miscanthus", "Panicum"),
    !is.na(mean),
    checked >= 0
  ) |>
  pull(mean)

if (length(sla_data) > 10 && exists("x") && exists("y")) {
  # Create plot comparing prior to histogram of data
  ggplot() +
    geom_histogram(
      data = data.frame(sla = sla_data),
      aes(x = sla, y = after_stat(density)),
      bins = 30, fill = "steelblue", alpha = 0.6
    ) +
    geom_line(
      data = data.frame(x = x, y = y),
      aes(x, y),
      color = "red", linewidth = 1, linetype = "dashed"
    ) +
    labs(
      title = "SLA: Prior Distribution vs. Observed Data",
      subtitle = "Red dashed = prior, Blue = observed data (Miscanthus + Panicum)",
      x = "SLA (m2/kg)",
      y = "Density"
    ) +
    xlim(0, 80)
}
```

## Extracting Data for a PFT

A common workflow: get all trait data for species in a given PFT.
```{r}
# Function to get trait data for a PFT
get_pft_data <- function(pft_name, trait_name = NULL) {
  # Find PFT id
  pft_id <- pfts |> 
    filter(grepl(pft_name, name, ignore.case = TRUE)) |>
    pull(id)
  
  if (length(pft_id) == 0) {
    warning("PFT not found: ", pft_name)
    return(NULL)
  }
  
  # Get species in this PFT
  spp_ids <- pfts_species |>
    filter(pft_id %in% !!pft_id) |>
    pull(specie_id)
  
  # Filter traitsview
  result <- traitsview |>
    filter(species_id %in% spp_ids)
  
  if (!is.null(trait_name)) {
    result <- result |> filter(trait == trait_name)
  }
  
  result
}

# Example usage
willow_sla <- get_pft_data("salix", "SLA")
if (!is.null(willow_sla) && nrow(willow_sla) > 0) {
  summary(willow_sla$mean)
}
```

## Session Info
```{r}
sessionInfo()
```