---
title: "Working with Plant Functional Types and Priors"
vignette: >
  %\VignetteIndexEntry{Working with Plant Functional Types and Priors}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(eval = requireNamespace("betydata", quietly = TRUE))
```

::: {.callout-note}
## What you will learn

- How Plant Functional Types (PFTs) organize species for ecosystem modeling
- How prior distributions encode parameter knowledge before Bayesian inference
- How to visualize priors and compare them to observed trait data
:::

## Background {#sec-background}

### Plant Functional Types

Ecosystem models simulate how plants exchange carbon, water, and energy with the atmosphere. These models require parameter values for photosynthesis rates, leaf properties, allocation fractions, and other ecophysiological quantities. Rather than parameterizing every species individually, models group species into **Plant Functional Types (PFTs)** -- categories like "temperate deciduous trees" or "C4 grasses" -- and estimate shared parameters for each group.

### Bayesian Parameterization

Bayesian inference is used to estimate model parameters. The workflow is:

1. **Prior distribution** $p(\theta)$: Encodes existing knowledge about a parameter $\theta$ before looking at data (e.g. "SLA is around 20 m2/kg with considerable uncertainty")
2. **Likelihood** $p(D|\theta)$: How probable the observed trait data $D$ is given parameter values
3. **Posterior distribution** $p(\theta|D)$: Updated belief after combining prior and data via Bayes' theorem:

$$
p(\theta | D) \propto p(D | \theta) \cdot p(\theta)
$$ {#eq-bayes}

The tables in `betydata` provide the ingredients for this workflow: `priors` stores prior distributions, `traitsview` stores the observed data, and `pfts`/`pfts_species`/`pfts_priors` define which priors and species belong to each PFT.

## Setup

```{r}
library(betydata)
library(dplyr)
library(ggplot2)
```

## Understanding PFTs {#sec-pfts}

### Available PFTs

```{r}
nrow(pfts)

pfts |>
  select(id, name) |>
  head(20)
```

### Species Membership

Each PFT contains species assigned by researchers. The `pfts_species` junction table maps PFTs to species:

```{r}
species_per_pft <- pfts_species |>
  count(pft_id, name = "n_species")

summary(species_per_pft$n_species)
```

```{r}
miscanthus_pft <- pfts |>
  filter(grepl("miscanthus", name, ignore.case = TRUE))

if (nrow(miscanthus_pft) > 0) {
  miscanthus_species <- pfts_species |>
    filter(pft_id %in% miscanthus_pft$id) |>
    left_join(
      species |> select(id, scientificname),
      by = c("specie_id" = "id")
    )
  miscanthus_species
}
```

## Working with Priors {#sec-priors}

### Prior Distribution Structure

Each row in the `priors` table specifies a probability distribution for a model parameter. The key fields are:

| Field         | Description                                      |
|---------------|--------------------------------------------------|
| `variable_id` | Which trait/parameter this prior describes        |
| `distn`       | Distribution family (e.g. `norm`, `gamma`, `lnorm`) |
| `parama`      | First parameter (mean, shape, etc.)               |
| `paramb`      | Second parameter (sd, rate, etc.)                 |

```{r}
#| label: tbl-distn-counts
#| tbl-cap: "Distribution families used in BETYdb priors"
priors |>
  count(distn, sort = TRUE) |>
  knitr::kable()
```

```{r}
#| label: tbl-prior-variables
#| tbl-cap: "Top variables with prior distributions"
priors |>
  left_join(variables |> select(id, name), by = c("variable_id" = "id")) |>
  count(name, sort = TRUE) |>
  head(15) |>
  knitr::kable()
```

### Visualizing a Prior Distribution {#sec-viz-prior}

Consider a prior for Specific Leaf Area (SLA). If the prior is $\text{Gamma}(\alpha, \beta)$, then the density function is:

$$
f(x; \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha - 1} e^{-\beta x}, \quad x > 0
$$ {#eq-gamma}

```{r}
#| label: fig-sla-prior
#| fig-cap: "Prior distribution for Specific Leaf Area (SLA)"
#| fig-width: 7
#| fig-height: 5
sla_var_id <- variables |>
  filter(name == "SLA") |>
  pull(id)

sla_priors <- priors |>
  filter(variable_id %in% sla_var_id)

eval_prior_density <- function(p, x = NULL) {
  if (is.null(x)) {
    x <- switch(p$distn,
      "gamma" = seq(0, qgamma(0.999, p$parama, p$paramb), length.out = 1000),
      "norm" = seq(p$parama - 4 * p$paramb, p$parama + 4 * p$paramb, length.out = 1000),
      "lnorm" = seq(0.001, qlnorm(0.999, p$parama, p$paramb), length.out = 1000),
      "beta" = seq(0.001, 0.999, length.out = 1000),
      "unif" = seq(p$parama, p$paramb, length.out = 1000),
      "weibull" = seq(0, qweibull(0.999, p$parama, p$paramb), length.out = 1000),
      "exp" = seq(0, qexp(0.999, p$parama), length.out = 1000),
      seq(0, 100, length.out = 1000)
    )
  }

  y <- switch(p$distn,
    "gamma" = dgamma(x, shape = p$parama, rate = p$paramb),
    "norm" = dnorm(x, mean = p$parama, sd = p$paramb),
    "lnorm" = dlnorm(x, meanlog = p$parama, sdlog = p$paramb),
    "beta" = dbeta(x, shape1 = p$parama, shape2 = p$paramb),
    "unif" = dunif(x, min = p$parama, max = p$paramb),
    "weibull" = dweibull(x, shape = p$parama, scale = p$paramb),
    "exp" = dexp(x, rate = p$parama),
    rep(0, length(x))
  )

  list(x = x, y = y)
}

if (nrow(sla_priors) > 0) {
  p <- sla_priors[1, ]
  prior_result <- eval_prior_density(p)
  prior_x <- prior_result$x
  prior_y <- prior_result$y

  ggplot(data.frame(x = prior_x, y = prior_y), aes(x, y)) +
    geom_line(linewidth = 1, color = "#2c3e50") +
    labs(
      title = paste0("Prior: ", p$distn, "(", round(p$parama, 2), ", ", round(p$paramb, 2), ")"),
      x = "SLA (m2/kg)",
      y = "Density"
    ) +
    theme_minimal(base_size = 12)
}
```

## Linking PFTs to Priors {#sec-pft-priors}

The `pfts_priors` table connects PFTs to their associated prior distributions, completing the Bayesian parameterization setup:

$$
\text{PFT} \xrightarrow{\texttt{pfts\_priors}} \text{Prior}(\theta) \xrightarrow{\texttt{variable\_id}} \text{Trait}
$$

```{r}
priors_per_pft <- pfts_priors |>
  count(pft_id, name = "n_priors")

summary(priors_per_pft$n_priors)
```

```{r}
#| label: tbl-pft-priors
#| tbl-cap: "Prior distributions for a Miscanthus PFT"
if (exists("miscanthus_pft") && nrow(miscanthus_pft) > 0) {
  pft_id_val <- miscanthus_pft$id[1]

  pft_priors <- pfts_priors |>
    filter(pft_id == pft_id_val) |>
    left_join(priors, by = c("prior_id" = "id")) |>
    left_join(variables |> select(id, name, units), by = c("variable_id" = "id"))

  pft_priors |>
    select(name, units, distn, parama, paramb) |>
    knitr::kable(digits = 3)
}
```

## Comparing Prior to Data {#sec-prior-vs-data}

A key validation step in Bayesian parameterization: compare the prior distribution to observed data. If the prior assigns negligible probability to regions where data actually falls, the prior may need revision.

::: {.callout-tip}
## Prior-Data Conflict

When the prior and data disagree strongly, the posterior will be dominated by whichever has lower variance. A very tight prior that disagrees with data can effectively ignore the data -- a clear signal to revisit prior specification.
:::

```{r}
#| label: fig-prior-vs-data
#| fig-cap: "SLA: Prior distribution vs. observed data for bioenergy grasses"
#| fig-width: 7
#| fig-height: 5
sla_data <- traitsview |>
  filter(
    trait == "SLA",
    genus %in% c("Miscanthus", "Panicum"),
    !is.na(mean),
    checked >= 0
  ) |>
  pull(mean)

if (length(sla_data) > 10 && exists("prior_x") && exists("prior_y")) {
  ggplot() +
    geom_histogram(
      data = data.frame(sla = sla_data),
      aes(x = sla, y = after_stat(density)),
      bins = 30, fill = "steelblue", alpha = 0.6
    ) +
    geom_line(
      data = data.frame(x = prior_x, y = prior_y),
      aes(x, y),
      color = "#e74c3c", linewidth = 1, linetype = "dashed"
    ) +
    labs(
      title = "Prior vs. Observed SLA",
      subtitle = "Dashed red = prior distribution | Blue bars = observed data (Miscanthus + Panicum)",
      x = "SLA (m2/kg)",
      y = "Density"
    ) +
    xlim(0, 80) +
    theme_minimal(base_size = 12)
}
```

## Extracting Data for a PFT {#sec-extract}

A common workflow: retrieve all trait data for species belonging to a given PFT.

```{r}
get_pft_data <- function(pft_name, trait_name = NULL) {
  pft_ids <- pfts |>
    filter(grepl(pft_name, name, ignore.case = TRUE)) |>
    pull(id)

  if (length(pft_ids) == 0) {
    warning("PFT not found: ", pft_name)
    return(tibble())
  }

  spp_ids <- pfts_species |>
    filter(pft_id %in% pft_ids) |>
    pull(specie_id)

  result <- traitsview |>
    filter(species_id %in% spp_ids)

  if (!is.null(trait_name)) {
    result <- result |>
      filter(trait == trait_name)
  }

  result
}

willow_sla <- get_pft_data("salix", "SLA")
if (nrow(willow_sla) > 0) {
  summary(willow_sla$mean)
}
```

## Key Concepts Summary

| Concept | Table(s) | Description |
|---------|----------|-------------|
| PFT definition | `pfts` | Named groupings of ecologically similar species |
| Species membership | `pfts_species` | Which species belong to which PFT |
| Prior specification | `priors` | Distribution family + parameters for each trait |
| PFT-Prior link | `pfts_priors` | Which priors apply to which PFT |
| Observed data | `traitsview` | Actual measurements for Bayesian updating |

## References

- LeBauer, D. S., et al. (2018). BETYdb: a yield, trait, and ecosystem service database. *GCB Bioenergy*. [doi:10.1111/gcbb.12420](https://doi.org/10.1111/gcbb.12420)
- LeBauer, D. S., et al. (2013). Facilitating feedbacks between field measurements and ecosystem models. *Ecological Monographs*, 83(2), 133--154.
