---
title: "Offline Analogs of BETYdb SQL Queries"
subtitle: "Common Database Queries Using betydata"
author: "Akash B V"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Offline Analogs of BETYdb SQL Queries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

## Introduction

This vignette demonstrates how to perform common BETYdb queries using `betydata` instead of connecting to a PostgreSQL database.

## Setup
```{r load-packages-data}
library(betydata)
library(dplyr)

# Load primary datasets
data(traitsview)
data(species)
data(sites)
data(managements)
data(treatments)
data(managements_treatments)
```

## Example 1: Query Miscanthus Yield Data

### Original SQL
```sql
SELECT yields.id, yields.mean, yields.date, 
       sites.sitename, species.scientificname
FROM yields
  LEFT JOIN sites ON yields.site_id = sites.id
  LEFT JOIN species ON yields.specie_id = species.id
WHERE species.genus = 'Miscanthus';
```

### betydata Equivalent
```{r miscanthus-yields}
miscanthus_yields <- traitsview |>
  filter(
    genus == "Miscanthus",
    trait == "Ayield"
  ) |>
  select(id, mean, date, sitename, scientificname)

head(miscanthus_yields)
nrow(miscanthus_yields)
```

## Example 2: Yield Data with Experimental Treatments

### Original SQL Pattern
```sql
SELECT yields.*, treatments.name AS treatment_name,
       managements.mgmttype, managements.level
FROM yields
  LEFT JOIN treatments ON yields.treatment_id = treatments.id
  LEFT JOIN managements_treatments ON treatments.id = managements_treatments.treatment_id
  LEFT JOIN managements ON managements_treatments.management_id = managements.id
WHERE species.genus IN ('Miscanthus', 'Panicum');
```

### betydata Equivalent
```{r yields-with-treatments}
# Join managements with treatments
mgmt_treat <- managements_treatments |>
  left_join(
    managements |> select(id, mgmttype, level, units, date),
    by = c("management_id" = "id")
  )

# Grass yields with management info
grass_yields <- traitsview |>
  filter(
    genus %in% c("Miscanthus", "Panicum"),
    trait == "Ayield"
  ) |>
  left_join(mgmt_treat, by = "treatment_id", relationship = "many-to-many")

# Summarize by management type
grass_yields |>
  filter(!is.na(mgmttype)) |>
  count(genus, mgmttype, sort = TRUE)
```

## Example 3: Extract Nitrogen Fertilization Rates

### Original dbplyr Pattern
```r
nitrogen <- managements %>% 
  filter(mgmttype == "fertilizer_N_rate") %>%
  select(treatment_id, nrate = level)
```

### betydata Equivalent
```{r nitrogen-rates}
# Get nitrogen rates per treatment
nitrogen_rates <- managements |>
  filter(mgmttype %in% c("fertilizer_N", "fertilizer_N_rate")) |>
  left_join(
    managements_treatments |> select(management_id, treatment_id),
    by = c("id" = "management_id")
  ) |>
  select(treatment_id, nrate = level, units)

# Join with yield data
yields_with_n <- traitsview |>
  filter(
    trait == "Ayield",
    genus %in% c("Miscanthus", "Panicum")
  ) |>
  left_join(nitrogen_rates, by = "treatment_id", relationship = "many-to-many")

# Summary
yields_with_n |>
  filter(!is.na(nrate)) |>
  group_by(genus) |>
  summarise(
    n = n(),
    mean_N = mean(nrate, na.rm = TRUE),
    mean_yield = mean(mean, na.rm = TRUE),
    .groups = "drop"
  )
```

## Example 4: Site-Level Aggregation

### Common Use Case

Aggregate trait data by site for spatial analysis.
```{r site-aggregation}
site_summary <- traitsview |>
  filter(!is.na(lat), !is.na(lon)) |>
  group_by(site_id, sitename, lat, lon) |>
  summarise(
    n_records = n(),
    n_traits = n_distinct(trait),
    n_species = n_distinct(species_id),
    .groups = "drop"
  )

# Top 10 most data-rich sites
site_summary |>
  arrange(desc(n_records)) |>
  head(10)
```

## Example 5: Citation-Based Queries

### Find All Data from a Specific Author
```{r author-search}
# Find all LeBauer data
lebauer_data <- traitsview |>
  filter(grepl("LeBauer", author, ignore.case = TRUE))

lebauer_data |>
  count(trait, author, citation_year, sort = TRUE)
```

### Citations with Most Data
```{r top-citations}
data(citations)

citation_counts <- traitsview |>
  count(citation_id, author, citation_year, sort = TRUE) |>
  head(15)

citation_counts
```

## Example 6: Variable/Trait Lookups

### Get Units and Description for a Trait
```{r variable-lookup}
data(variables)

# Look up specific traits
variables |>
  filter(name %in% c("SLA", "Vcmax", "leaf_respiration_rate_m2")) |>
  select(name, units, description, min, max)
```

## Performance Comparison

Because `betydata` datasets are pre-loaded R data frames, queries are typically faster than database connections:
```{r performance-test}
# Time a complex filter
system.time({
  result <- traitsview |>
    filter(
      genus %in% c("Miscanthus", "Panicum", "Populus"),
      trait %in% c("SLA", "Vcmax", "Ayield"),
      checked == 1
    ) |>
    group_by(genus, trait) |>
    summarise(n = n(), mean = mean(mean, na.rm = TRUE), .groups = "drop")
})
```

## Session Info
```{r session-info}
sessionInfo()
```